---
title: "Introduction to Data Science"
subtitle: "Strings, text processing, and regexp"
date: "Fall 2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    #css: ["default","metropolis","metropolis-fonts","animate.css"]
    css: ["xaringan-themer.css","metropolis-fonts"]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
    # df_print: tibble
    # css: [default, metropolis, metropolis-fonts]
    # nature:
    #   highlightStyle: zenburn
    #   highlightLines: true
    #   countIncrementalSlides: false
--- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.retina = 3, fig.asp = 0.6, fig.align = 'center', out.width = "120%",message = FALSE)
options(htmltools.dir.version = FALSE, digits = 3, knitr.table.format = "html",tibble.print_min=6, tibble.width=70)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
xaringanExtra::use_tile_view()
xaringanExtra::use_scribble()
xaringanExtra::use_extra_styles(hover_code_line = TRUE)
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_tachyons()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
style_duo_accent(primary_color = "#035AA6", secondary_color = "#03A696",
   #title_slide_background_color = "#FFFFFF",
   #title_slide_text_color = "#006747",
   link_color = "#03A696",
   header_font_google = google_font("Josefin Sans"),
   title_slide_background_image = "belltower2.jpg",
   #title_slide_background_size = "600px",
   #title_slide_background_position = "bottom",
   text_font_google   = google_font("Montserrat", "300", "300i"),
   code_font_size = "0.8rem",
   code_font_family = "Fira Code",
   code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```

# Stringr

Strings processing in **R** can be done using the [stringr](https://stringr.tidyverse.org/) library.

For more documentation on **stringr**, see 

+ The [stringr vignette](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)

+ The [stringr cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf)

+ Chapter 14 of [R for data science](https://r4ds.had.co.nz/strings.html)

+ Chapter 24 of [Introduction to data science](https://rafalab.github.io/dsbook/string-processing.html)

---
# String basics

+ A string in **R** is created using either `'...'` or `"..."`
```{r}
a <- 'This is a string'; a
b <- "This is another string"; b
c <- 'And he said "Dude, where is my car?"'; c
```

---
class: middle
+ A string in **R** is a **character vector**, so it has length $1$.
```{r}
a <- "A string"
length(a)
b <- "A longer longer longer string"
length(b)
```

---
class: clear

+ Use `\` to quote special characters in a string. Use `\\` to get `\` in a string. Use `writeLines()` to get the **raw** string. 
```{r}
a <- "This string has a \t (tab)"; a
writeLines(a)
b <- "And this string has \n two lines"; b
writeLines(b)
c <- "And this string has a \\ (slash)"; c
writeLines(c)
```

---
# Some useful stringr functions

+ `str_length` for length of string(s)
```{r}
    library(stringr)
    x <- c("One fish, two fish, red fish, blue fish", 
           "Green eggs and ham", 
           "Oh say can you say")
    str_length(x)
```

---
class: middle

+ `str_c` for combining strings
```{r}
str_c("Alice", "Bob", "Charles")
str_c("Alice", "Bob", "Charles", sep = ",")
str_c("Alice", "Bob", "Charles", sep = "\t") %>% writeLines()
## str_c is vectorized
str_c(c("One", "Two", "Three"), c("Alice", "Bob", "Charles"), sep = " ") 
## vectorized with vector recycling 
str_c("One", c("Alice", "Bob", "Charles"), 
             c("Oat", "Almond", "Walnut"), sep = " ")
```

---
class: middle

+ The **collapse** argument for **str_c** is useful to turn a vector of strings into individual arguments.
```{r}
str_c(c("Alice","Bob","Charles"),sep=",")
str_c(c("Alice","Bob","Charles"), collapse = ",")
```
+ `str_glue` is useful for inserting **R** output into a string.
```{r}
f <- function(x){
  2^x
}
x <- c(2,3,4)
str_glue("The value of two to the power of {x} is {f(x)}")
```
---
class: middle
+ `str_pad` and `str_trunc` for padding and truncating strings.
```{r}
x <- as.character(c(5,20, 80, 320, 1280, 27607, 111111))
x %>% str_trunc(5) %>% str_pad(5, side = "left", pad = "0")
x %>% str_trunc(5, ellipsis = "") %>% str_pad(5, side = "left", pad = "0")
```
+ `str_sub` to extract substrings from a string.  
```{r}
x <- c("One fish", "Two ship", "Three cats", "Four beeps")
str_sub(x, 2, 5) ## Extract second to fifth character in each string
str_sub(x, -5, -2) ## Extract second to fifth character (backwards)
str_sub(x, 2, 5) <- "Haha" ## Assigning things to a substring
x
```

---
class: middle

+ `str_sort` and `str_order` to order/sort the characters in a string.
```{r}
x <- c("winner winner", "chicken dinner", 
       "cherry blossom", "eager beaver") 
x %>% str_sort()
x %>% str_sort(locale = "haw") ## Hawaiian's don't speak English!
```

+ To sort the letters in a string, you will have to first split the string using `str_split`
```{r}
## Warning. This code chunk is a quick hack. 
## In practice we will try to write something more elegant.
sapply(lapply(str_split(x, ""), str_sort), function(z) str_c(z, collapse = ""))
```