---
title: "Introduction to Data Science"
subtitle: "Strings, text processing, and regexp"
date: "Fall 2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    #css: ["default","metropolis","metropolis-fonts","animate.css"]
    css: ["xaringan-themer.css","metropolis-fonts"]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
    # df_print: tibble
    # css: [default, metropolis, metropolis-fonts]
    # nature:
    #   highlightStyle: zenburn
    #   highlightLines: true
    #   countIncrementalSlides: false
--- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.retina = 3, fig.asp = 0.6, fig.align = 'center', out.width = "120%",message = FALSE)
options(htmltools.dir.version = FALSE, digits = 3, knitr.table.format = "html",tibble.print_min=6, tibble.width=70)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
xaringanExtra::use_tile_view()
xaringanExtra::use_scribble()
xaringanExtra::use_extra_styles(hover_code_line = TRUE)
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_tachyons()
xaringanExtra::use_animate_css()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
style_duo_accent(primary_color = "#035AA6", secondary_color = "#03A696",
   #title_slide_background_color = "#FFFFFF",
   #title_slide_text_color = "#006747",
   link_color = "#03A696",
   header_font_google = google_font("Josefin Sans"),
   title_slide_background_image = "belltower2.jpg",
   #title_slide_background_size = "600px",
   #title_slide_background_position = "bottom",
   text_font_google   = google_font("Montserrat", "300", "300i"),
   code_font_size = "0.8rem",
   code_font_family = "Fira Code",
   code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```

# Stringr

Strings processing in **R** can be done using the [stringr](https://stringr.tidyverse.org/) library.

For more documentation on **stringr**, see 

+ The [stringr vignette](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)

+ The [stringr cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf)

+ Chapter 14 of [R for data science](https://r4ds.had.co.nz/strings.html)

+ Chapter 24 of [Introduction to data science](https://rafalab.github.io/dsbook/string-processing.html)

---
# String basics
+ A string in **R** is created using either `'...'` or `"..."`
```{r}
a <- 'This is a string'; a
b <- "This is another string"; b
c <- 'And he said "Dude, where is my car?"'; c
```

---
class: middle
+ A string in **R** is a **character vector**, so it has length $1$.
```{r}
a <- "A string"
length(a)
b <- "A longer longer longer string"
length(b)
```

---
class: clear

+ Use `\` to quote special characters in a string. Use `\\` to get `\` in a string. Use `writeLines()` to get the **raw** string. 
```{r}
a <- "This string has a \t (tab)"; a
writeLines(a)
b <- "And this string has \n two lines"; b
writeLines(b)
c <- "And this string has a \\ (slash)"; c
writeLines(c)
```

---
# Basic stringr functions

+ `str_length` for length of string(s)
```{r}
    library(stringr)
    x <- c("One fish, two fish, red fish, blue fish", 
           "Green eggs and ham", 
           "Oh say can you say")
    str_length(x)
```

+ `str_c` for combining strings
```{r}
str_c("Alice", "Bob", "Charles")
str_c("Alice", "Bob", "Charles", sep = ",")
str_c("Alice", "Bob", "Charles", sep = "\t") %>% writeLines()
```
---
class: middle
+ stringr functions are **vectorized** by default.
```{r}
str_c(c("One", "Two", "Three"), c("Alice", "Bob", "Charles"), sep = " ") 
## vectorized with vector recycling 
str_c("One", c("Alice", "Bob", "Charles"), 
             c("Oat", "Almond", "Walnut"), sep = " ")
```

+ The **collapse** argument for **str_c** is useful to turn a vector of strings into individual arguments.
```{r}
str_c(c("Alice","Bob","Charles"),sep=",")
str_c(c("Alice","Bob","Charles"), collapse = ",")
```
---
class: middle
+ `str_glue` is useful for inserting **R** output into a string.
```{r}
f <- function(x){
  2^x
}
x <- c(2,3,4)
str_glue("The value of two to the power of {x} is {f(x)}")
```

+ `str_pad` and `str_trunc` for padding and truncating strings. 
`str_trim` for trimming white spaces from start and end of a string.
```{r}
x <- as.character(c(5,20, 80, 320, 1280, 27607, 111111))
x %>% str_trunc(5) %>% str_pad(5, side = "left", pad = "0")
x %>% str_trunc(5, ellipsis = "") %>% str_pad(5, side = "left", pad = "0")
```
---
class: middle 
+ `str_sub` to extract substrings from a string.  
```{r}
x <- c("One fish", "Two ship", "Three cats", "Four beeps")
str_sub(x, 2, 5) ## Extract second to fifth character in each string
str_sub(x, -5, -2) ## Extract second to fifth character (backwards)
str_sub(x, 2, 5) <- "Haha" ## Assigning things to a substring
x
```
---
class: middle
+ `str_sort` and `str_order` to order/sort the characters in a string.
```{r}
x <- c("winner winner", "chicken dinner", 
       "cherry blossom", "eager beaver") 
x %>% str_sort()
x %>% str_sort(locale = "haw") ## Hawaiian's don't speak English!
```

+ To sort the letters in a string, you will have to first split the string using `str_split`.
```{r}
## This code chunk is just a quick hack. 
sapply(lapply(str_split(x, ""), str_sort), function(z) str_c(z, collapse = ""))
```
---
# Pattern matching in strings

Most uses of **stringr** is to search for and match patterns in string. Some of the most widely used functions are

+ `str_detect`, `str_which` and `str_subset` to detect presence of a pattern in a character vector (collection of strings).
```{r}
x <- c("apple", "pear", "banana", "orange")
str_detect(x, "p")
str_which(x, "p")
str_subset(x, "an")
```

---
class: clear

+ `str_count` for counting the number of matches and `str_locate` for the (starting and ending) location index of the (first) match. See also `str_locate_all`.

```{r}
x
str_count(x, "p")
str_count(x, "ana")
str_locate(x, "ana")
```

---
class: clear

+ `str_replace` and `str_replace_all` for search and replace.
```{r}
fruit %>% head ## A character vector of fruits contained in the stringr package.
fruit %>% str_replace("a", "z") %>% head
fruit %>% str_replace_all(c("a" = "z", "b" = "y")) %>% head
```

+ `str_extract` and `str_match` to extract matching patterns from a string. These functions are, however, only useful when used in combination with **regular expressions**.
```{r}
fruit %>% str_extract("pp") %>% head
```
---
# Pop Quiz

Q1. What is a regular expression ? 

--

A1. Anything that is recognizable by a finite state automata.

--

Q2. But what is a finite state automata ?

--

A2. Anything whose set of recognizable strings are generated by a regular expression.

---
# Regular expressions

Regexps provide a powerful tool for describing patterns in strings. 
```{r}
## which fruits' names contain at least two consecutive vowels ?
fruit %>% str_subset("[aeiou]{2,}") %>% head(4)
## which fruits' names contain the letter y but is not a berry ?
fruit %>% str_subset("[^y]$") %>% str_subset("y") %>% head(4)
## which fruits' names has the same letter appearing consecutively 
fruit %>% str_subset("(.)\\1") %>% head(4)
## which fruits' names has the same two letters appearing consecutively 
fruit %>% str_subset("(..)\\1") %>% head(4)
```

---
class: middle
We can now see the difference between `str_match` and `str_extract`
```{r}
regexp1 <- "(..)\\1"
fruit %>% str_extract(regexp1) %>% head(4)
fruit %>% str_match(regexp1) %>% head(4)
bands <- c("ABBA", "Pink Floyd", "Nirvana", "Eagles")
regexp2 <- "(.)(.)\\2\\1"
bands %>% str_extract(regexp2)
bands %>% str_match(regexp2)
```

---
class: clear

For more on regular expressions, see 

+ Section 14.3 of [R for Data Science](https://r4ds.had.co.nz/strings.html)

+ The regexp [cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf)

The most important things to remember are

+ `^` and `$` to anchor the pattern at the beginning or at the end of the string

+ `.` matches any character, `\d` matches any digits, `\s` matches any whitespace character.

+ `[abc]` matches `a`, `b` or `c` while `[^abc]` matches **not** `a`, `b` or `c`

+ `?` to specify that a pattern is matched $0$ or $1$ time, `+` to specify that a pattern is matched at least $1$ time and `*` to specify that a pattern is matched $0$ or more times.

+ `\\1`, `\\2`, for back references
---
# Example 1: Macbeth
We will follow the presentation from Chapter 19 of the book [Modern Data Science with R](https://mdsr-book.github.io/mdsr2e/ch-text.html#regular-expressions-using-macbeth)

```{r}
library(mdsr)
library(dplyr)
data(Macbeth_raw)
str(Macbeth_raw)
```

The data is representated as a single string. We first split this string into multiple lines.
```{r}
macbeth_split <- Macbeth_raw %>% str_split("\r\n")
macbeth <- macbeth_split[[1]]
length(macbeth)
```

---
class: middle

The data is now in a much simpler format.
```{r}
macbeth[300:310]
```

---
class: middle
We can now count how many times a particular character speaks.
```{r}
thus_spoke_macbeth <- macbeth %>% str_subset("^\\s+MACBETH")
length(thus_spoke_macbeth)
head(thus_spoke_macbeth, 4)
macbeth %>% str_subset("^\\s+MAC(BETH|DUFF)") %>% length()
```

---
class: middle

The four most important characters in *Macbeth* are Macbeth himself, Lady Macbeth (his wife), Banquo (his friend) and Duncan (the king). Let us now plot how frequent these characters speak during the play.

```{r}
macbeth_chars <- tribble(
  ~name, ~regexp,
  "MACBETH", "^\\s+MACBETH",
  "LADY MACBETH", "^\\s+LADY MACBETH",
  "BANQUO", "^\\s+BANQUO",
  "DUNCAN", "^\\s+DUNCAN") %>% 
  mutate(speaks = purrr::map(regexp, str_detect, string = macbeth)) 
macbeth_chars ## Oh hey, list columns!!
str(macbeth_chars$speaks[[1]]) 
```

---
class: middle

Let us now **tidy** up this data by first unnesting the list column.
```{r}
library(tidyr)
macbeth_chars <- macbeth_chars %>% unnest(speaks)
macbeth_chars
```

Next, let us add in the line numbers for each sentence and also remove the unimportant header lines.
```{r}
macbeth_chars <- macbeth_chars %>% 
  mutate(lineno = rep(1:length(macbeth),4)) %>% 
  filter(lineno > 218, lineno < 3172)
glimpse(macbeth_chars)
```

---
class: middle

We now visualize the frequency at which a character speaks in the play.
However, since `speaks` only take on the $\{0,1\}$ values, it is not clear how to visualize this information. 
The following is one, somewhat hackish, approach.
```{r out.width="60%", warning = FALSE}
library(ggplot2)
p <- macbeth_chars %>% ggplot(aes(x = lineno, y = as.numeric(speaks))) + 
  geom_smooth(aes(color = name), method = "loess", se = FALSE, span = 0.4) + 
  xlab("Line number") + ylab("Frequency of speaking") + ylim(c(0,NA))
p
```

---
class: middle
Finally, we can add some information about which part of the play each line corresponds to.
```{r out.width="60%", fig.align = "center", warning = FALSE}
acts <- tibble(
  lineno = str_which(macbeth, "^ACT [I|V]+"), 
  line_text = str_subset(macbeth, "^ACT [I|V]+"),
  labels = str_extract(line_text, "^ACT [I|V]+"))
p <- p + geom_vline( data = acts, aes(xintercept = lineno), 
                     color = "darkgray", lty = 3) + 
         geom_text(data = acts, aes(y = 0.085, label = labels), 
                   hjust = "left", color = "darkgray") 
p
```

---
# Example: Self-reported heights
Our next example is from Chapter 24.3 of [Introduction to Data Science](https://rafalab.github.io/dsbook/string-processing.html). A total of $1095$ students were asked to report their heights in inches, with quite a few problematic entries.
```{r}
library(dslabs)
data(reported_heights)
reported_heights <- reported_heights %>%
  mutate(new_height = as.numeric(height))
reported_heights %>% filter(is.na(new_height)) %>% summarise(n())
```
---
A random sample of all the problematic entries are
```{r}
not_inches <- function(x, smallest = 50, tallest = 84){
  inches <- suppressWarnings(as.numeric(x))
  ind <- !is.na(inches) & (inches >= smallest & inches <= tallest) 
  !ind
}
## First remove extraneous spaces
reported_heights <- reported_heights %>% mutate(height = str_trim(height, side = "both")) 
reported_heights %>% filter(not_inches(height)) %>% 
  pull(height) %>% sample(60)
```

---
We first try to do some search and replace to convert the simple issues like `5' 4` or `5' 3"` into inches.
We start by creating a pattern to search for these issues. 
```{r}
problems <- reported_heights %>% 
  filter(not_inches(height)) %>% pull(height) 
length(problems) 
pattern1 <- "^([4-7])'\\s*(\\d+)\"?" 
problems %>% str_subset(pattern1)
```
---
class: clear
We then extract the feet and inches from entries matching this pattern.
```{r}
problems %>% str_subset(pattern1) %>% str_match(pattern1) %>% head
```
We now fixed these problematic entries.
```{r}
first_issue <- problems %>% str_subset(pattern1) %>% str_match(pattern1)
first_issue_new_height <- as.character(
  as.numeric(first_issue[,2])*12 + as.numeric(first_issue[,3]))
problems[str_which(problems, pattern1)] <- first_issue_new_height
length(problems[not_inches(problems)])
```

---
We next consider the issue of heights reported in centimeters.
```{r}
pattern2 <- "^([12]\\d{2})\\s*(cm)?$"
second_issue <- problems %>% str_subset(pattern2)
second_issue
## Remove cm and trim white space
second_issue <- str_replace(second_issue,"cm","") %>% str_trim(side = "right")
## Convert from cm to inches
second_issue_new_height <- as.character(round(as.numeric(second_issue)/2.54))
problems[str_which(problems, pattern2)] <- second_issue_new_height
length(problems[not_inches(problems)])
```

---
Let us now take a look at a sample of the remaining issues
```{r}
problems[not_inches(problems)] %>% sample(60)
```
---
The third issue could be adding `ft` or `feet` or `inches` to the reported heights.
We can easily fix this by first doing a `str_replace` and then using our `pattern1` above.
```{r}
str_subset(problems,"(ft)|(feet)|(inches)|")
problems <- str_replace(problems,"(ft)|(feet)|(foot)","'") 
problems <- str_replace(problems,"(inches)","\"")
problems <- str_replace_all(problems,"\\s","") ## Also remove spaces
problems <- str_replace(problems,"and","")
pattern1 <- "^([4-7])'\\s*([\\d\\.]+)\"?"
first_issue <- problems %>% str_subset(pattern1) %>% str_match(pattern1)
first_issue_new_height <- as.character(
  round(as.numeric(first_issue[,2])*12 + as.numeric(first_issue[,3])))
problems[str_which(problems, pattern1)] <- first_issue_new_height
length(problems[not_inches(problems)])
```

---
The fourth issue could be the use of `.` or `,` to separate between feet and inches.
This can be fixed using a `str_match`
```{r}
pattern3 <- "^([4-7])[\\.,]([\\d\\.]+)"
fourth_issue <- problems %>% str_subset(pattern3) %>% str_match(pattern3)
fourth_issue_new_height <- as.character(
  round(as.numeric(fourth_issue[,2])*12 + as.numeric(fourth_issue[,3])))
problems[str_which(problems, pattern3)] <- fourth_issue_new_height
length(problems[not_inches(problems)])
```

---
The fifth issue could be that people who are 5 feet or 6 feet will just write `5` or `6`. 
This is easy enough to fix.
```{r}
pattern4 <- "^[4-7]'?$"
fifth_issue <- str_subset(problems,pattern4)
fifth_issue <- str_replace(fifth_issue,"'","")
problems[str_which(problems,pattern4)] <- 
  as.character(as.numeric(fifth_issue)*12)
problems[not_inches(problems)]
```

---
Finally we see that there are a few remaining issues such as (1) using `,` or `.` in centimeters and (2) including `"` in the inches measurement. These are also relatively easy to fix.
```{r}
problems <- problems %>% str_replace("\"$","")
pattern2 <- "^([12])[\\,\\.]?(\\d{1,2})$"
sixth_issue <- problems %>% str_subset(pattern2) %>% str_match(pattern2)
## First get the centimeters
sixth_issue_heights_cm <- as.numeric(sixth_issue[,2])*100 + 
  as.numeric(str_pad(sixth_issue[,3],width=2,pad="0",side = "right"))
## Convert from cm to inches
sixth_issue_new_height <- as.character(round(sixth_issue_heights_cm/2.54))
problems[str_which(problems, pattern2)] <- sixth_issue_new_height
problems[not_inches(problems)]
```
The remaining issues are unlikely to be fixable automatically. We will have to ignore them.
---
We update our data frame to have a new height column.
```{r}
reported_heights <- reported_heights %>% mutate(new_height = height) %>% 
  filter(not_inches(height)) %>% mutate(new_height = problems) 
reported_heights %>% filter(!not_inches(new_height)) %>% head(10)
```
--
Q. What did we learned from this example ?

A. It was quite tedious and boring. Regrettably being a data scientist is not always swashbuckling fun. Luckily enough, it pays well.
