---
title: "Clustering exaples"
output: html_document
---

# Example 1: Hierarchial clustering of Toyota cars

We first read in the data.
```{r cache = TRUE, message = FALSE, warning = FALSE}
library(dplyr)
library(readxl)
filename <- fs::dir_ls("16data", regexp = "public\\.xlsx") %>% head(1)
library(janitor)
library(tibble)
cars <- read_excel(filename) %>% 
  janitor::clean_names() %>%
  rename(make = mfr_name, model = carline, displacement = eng_displ,
    city_mpg = city_fe_guide_conventional_fuel,
    hwy_mpg = hwy_fe_guide_conventional_fuel) %>%
  select(make, model, displacement, number_cyl, number_gears, city_mpg, hwy_mpg) %>%
  distinct(model, .keep_all = TRUE) %>% 
  filter(make == "Toyota") %>%
  column_to_rownames(var = "model")
```

There are in total `r nrow(cars)` Toyotal cars. A snippet of the data is as follows.
```{r}
head(cars)
```

We next compute a dissimilarity/distance between the cars.
```{r}
cars_scaled <- scale(dplyr::select(cars, -make), center = TRUE, scale = TRUE)
toyota_dist <- dist(cars_scaled) 
as.matrix(toyota_dist)[1:5,1:5] ## Distance between the first five cars
```

```{r fig.retina = 3, fig.asp = 0.8, out.width="120%"}
cars_hclust <- hclust(toyota_dist, method = "complete") ## Complete linkage
plot(cars_hclust, cex = 0.5, hang = -1, xlab = "", sub = "", 
     main = "Hierarchical clustering of Toyota cars (complete linkage)")
```

It is hard to see the labels of the previous hierarchical clustering as they are rotated.
A better way to visualize this dendogram is to draw the tree horizontally. We can do this via the **ape** package for analyzing phylogenetic trees.

```{r fig.retina = 3, fig.asp = 1.2,out.width="120%"}
library(ape)
cars_hclust <- as.phylo(cars_hclust)
plot(cars_hclust, cex = 0.5, label.offset = 0.2, main = "Complete Linkage")
```

We now compare the complete linkage hierarchical clustering with the single linkage and the average linkage clustering.

```{r fig.retina = 3, fig.asp = 1.2, out.width = "120%"}
cars_hclust_single <- as.phylo(hclust(toyota_dist, method = "single")) ## Single linkage
cars_hclust_avg <- as.phylo(hclust(toyota_dist, method = "average")) ## Average linkage
plot(cars_hclust_single, cex = 0.5, label.offset = 0.2, main = "Single Linkage")
plot(cars_hclust_avg, cex = 0.5, label.offset = 0.2, main = "Average Linkage")
```

We see that the clusterings differ substantially. In particular, the single linkage clustering has much longer paths from the root to the leaf node.

We now try to cut the complete linkage tree into five clusters.
```{r}
cars_hclust <- hclust(toyota_dist, method = "complete") ## Complete linkage
memb <- cutree(cars_hclust, k = 5)
head(memb)
table(memb)
```