---
title: "CSC/ST 442 (Fall 2021): Assignment 2"
urlcolor: blue
output: 
  
  pdf_document:
    df_print: tibble
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3, knitr.table.format = "html",tibble.print_min=6)
```

## Instruction
This assignment consists of $3$ problems. The assignment is due on 
**Thursday, September 16** at 11:59pm EDT. Please submit your assignment electronically through the moodle webpage. You are encouraged (but not required) to use RMarkdown to write up your homework solution. To start using Rmarkdown read

* Section 40.2 of  [Introduction to Data Science](https://rafalab.github.io/dsbook/reproducible-projects-with-rstudio-and-r-markdown.html)
* the [RStudio tutorial](https://rmarkdown.rstudio.com/lesson-1.html) 
* the [Rmarkdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).



## Problem 1 (30pts)
This problem uses the *flights* and *weather* dataset from the **nycflights13** library. A snippet of the data is as follows.
```{r}
library(nycflights13)
flights
weather
```

Using these two datasets, answer the following question.

1. (5pts) Visualize the departure times of cancelled versus non-cancelled flights. 
2. (5pts)  Draw a boxplot of the temperatures, grouped by months.
3. (10pts) Visualize the proportion of cancelled flights each day against the average daily temperature, grouped by the airports of origin (EWR, JFK, LGA). 

Hint: you need to create a new column that maps the month and day to a number corresonding to the number of days since January 1. Try this
```{r}
as.numeric(as.Date(paste(2013,2,24,sep="/"),format="%Y/%m/%d") - 
             as.Date("2013/01/01",format="%Y/%m/%d"))
```
A more elegant handling of dates and time is available via the [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) library, but the above is sufficient for our current purpose. 

4. (10pts) Visualize the arrival delay for flights to the top twenty most popular destinations, grouped by destinations. 

## Problem 2 (20pts)
This problem uses the **us_contagious_diseases** dataset from the [dslabs](https://cran.r-project.org/web/packages/dslabs/index.html) library. This library accompanies the book [Introduction to Data Science](https://rafalab.github.io/dsbook/) by R. Irizarry. A snippet of the data is as follows.
```{r}
library(dslabs)
data(us_contagious_diseases)
us_contagious_diseases
```
Each row of the data frame `us_contagious_diseases` record the **yearly** total number of cases of a specific disease (in each of the $50$ states) during the period $1928$ to $2011$. For example the first row say that there was $321$ reported cases of Hepatitis A in Alabama during the year $1966$. 

Using this data, answer the four questions in [Section 10.15](https://rafalab.github.io/dsbook/data-visualization-principles.html#exercises-20) of the book Introduction to Data Science. You might want to read through the code examples in [Section 10.14](https://rafalab.github.io/dsbook/data-visualization-principles.html#vaccines) of that book. 



## Problem 3 (25pts)
This problem uses the obesity data set from the CDC. The original data is in Excel format and we had uploaded a copy of this file onto Moodle. First download the data onto your desktop/laptop work space
and extract the data as follows. 
```{r message = FALSE, warning = FALSE,error=TRUE}
library(readxl) ## install.packages("readxl") if the readxl library is missing.
fname <- "obesity_data.xlsx"
wrkbk <- read_excel(fname)
obesity_2012 <- setNames(wrkbk[-1, c(2, 61)], c("fips", "pct"))
obesity_2012$pct <- as.numeric(obesity_2012$pct) / 100
obesity_2012
```

We next load the socviz library and get access to the boundaries line for the US map.
```{r}
library(socviz) ## install.packages("socviz") if the socviz library is missing.
county_map
county_data
```

Combining the $\mathtt{obesity\_2012}$ data frame and either the $\mathtt{county\_map}$ or $\mathtt{county\_data}$ data, generate a visualization of the US Obesity Rate by County. Using the $\mathtt{county\_data}$, find the variables that are "correlated" with obesity rates. 