---
title: "Sample questions"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```

# Part 1 (Wrangling data with dplyr)

For the next several questions, use the following dataset.

```{r}
library(gapminder)
gapminder
```

+ Q1: What is the most accurate way to compute the average life expectancy for each continent in 1952 ?
```{r eval = FALSE}
## (a) 
gapminder %>% filter(year == 1952) %>% summarise(avg_life_Exp = mean(lifeExp))
    
## (b)
gapminder %>% filter(year == 1952) %>% mutate(total_life = lifeExp*pop) %>% 
                  summarise(avg_life_Exp = mean(total_life))

## (c)
gapminder %>% group_by(continent) %>% filter(year == 1952) %>% 
  mutate(total_life = lifeExp*pop) %>%
  summarise(total_life = sum(total_life), total_pop = sum(pop)) %>% 
  mutate(avg_life_Exp = total_life/total_pop)

## (d)
gapminder %>% filter(year == 1952) %>% group_by(continent) %>% 
  summarise(avg_life_Exp = mean(lifeExp))
```

+ Q2: How will you compute, for each year, the standardized gdpPercap for each country 
(standardized by the continent the country belongs to) ?

```{r error = TRUE, message = FALSE, eval = FALSE}
library(dplyr)

## (a) 
gapminder %>% group_by(year) %>% mutate(std_gdp = scale(gdpPercap))

## (b)
gapminder %>% group_by(continent, year) %>% mutate(std_gdp = scale(gdpPercap))

## (c)
gapminder %>% group_by(continent, year) %>% summarise(std_gdp = scale(gdpPercap))

## (d)
gapminder %>% group_by(year) %>% summarise(std_gdp = scale(gdpPercap))
```

+ Q3: The following table is obtained by splitting the **gapminder** data into $3$ groups based on the logarithm of gdpPercap in $1952$ (these buckets corresponds roughly the "undeveloped", "developing", and "developed" countries at that time) and then find the country with the sharpest population increase between $1952$ and $2007$, **grouped by the original development status in $1952$**. Which of the following code will reproduce this table ?

```{r echo = FALSE, message = FALSE}
library(dplyr)
gapminder_1952 <- gapminder %>% filter(year == 1952) %>% mutate(develop = ntile(log(gdpPercap),3)) %>% select(country, develop)
gapminder %>% left_join(gapminder_1952) %>% group_by(country) %>% 
  mutate(pop_growth = pop[year == 2007]/pop[year == 1952]) %>% 
  ungroup() %>% group_by(develop) %>% filter(pop_growth == max(pop_growth), year %in% c(1952, 2007))
```

```{r eval = FALSE,error = TRUE}

## (a)
gapminder %>% mutate(develop = ntile(log(gdpPercap), 3)) %>% group_by(country) %>% 
  mutate(pop_growth = pop[year == 2007]/pop[year == 1952]) %>% 
  ungroup() %>% group_by(develop) %>% 
  filter(pop_growth == max(pop_growth), year %in% c(1952, 2007))

## (b)
gapminder_1952 <- gapminder %>% filter(year == 1952) %>% 
  mutate(develop = ntile(log(gdpPercap),3)) %>% 
  select(country, develop)
gapminder %>% left_join(gapminder_1952) %>% group_by(country) %>% 
  mutate(pop_growth = pop[year == 2007]/pop[year == 1952]) %>% 
  ungroup() %>% group_by(develop) %>% 
  filter(pop_growth == max(pop_growth), year %in% c(1952, 2007))

## (c)
gapminder_1952 <- gapminder %>% filter(year == 1952) %>% 
  mutate(develop = ntile(log(gdpPercap),3)) 
gapminder %>% left_join(gapminder_1952) %>% group_by(country) %>% 
  mutate(pop_growth = pop[year == 2007]/pop[year == 1952]) %>% 
  filter(pop_growth == max(pop_growth), year %in% c(1952, 2007))

## (d)
gapminder %>% mutate(develop = ntile(log(gdpPercap), 3)) %>% 
  group_by(country,develop) %>% 
  mutate(pop_growth = pop[year == 2007]/pop[year == 1952]) %>% 
  filter(pop_growth == max(pop_growth), year %in% c(1952, 2007))
```

# Part 2 (Base R datastructures and automation)
Consider the following list
```{r}
list1 <- list(a = list(1,2), b = list(3,3), c = list("a","b","c","d"))
```

+ Q1. What is the length of $\mathtt{list1}$ ?
$$\mathrm{(a)} 4; \quad \mathrm{(b)}  2; \quad  \mathrm{(c)} 5; \quad \mathrm{(d)} 3$$
+ Q2. What is the length of $\mathtt{list1[2]}$ ?
$$\mathrm{(a)} 1; \quad \mathrm{(b)}  2; \quad  \mathrm{(c)} 3; \quad \mathrm{(d)} 4$$

+ Q3. What is the result of the following code chunk ?
```{r eval = FALSE}
library(magrittr) ## This allows the use of the %>% operator
list1 %>% lapply(length) %>% sapply(unique)
```
$$\mathrm{(a)} [1,2,4]; \quad \mathrm{(b)}  [2,2,4]; \quad  \mathrm{(c)} [2,1,4]; \quad \mathrm{(d)} [2,4]$$

+ Q4: What is the relationship between a list and a vector in **R** ?
  + (a) A list is a special type of vector.
  + (b) A vector is a special kind of list.
  + (c) A vector and a list are distinct data types.
  + (d) The intersection of a list and a vector is the **NULL** object.

# Part 3 (Tidying Data)
Consider the following table, call it $\mathtt{table1}$.
```{r echo = FALSE, message = FALSE}
library(tidyverse)
table1 <- tribble(
  ~country, ~year, ~cases, ~population, ~expense,
  "Afghanistan", 1999, 745, 19987071, 100,
  "Afghanistan", 2000, 2666, 20595360, 150,
  "Brazil", 1999, 37737, 172006362, 125,
  "Brazil", 2000, 80488, 174504898, 200,
  "China", 1999, 212258, 1272915272, 250,
  "China", 2000, 213766, 1280428583, 175,
)

table2 <- gather(table1, key = "type", value = "value", cases, population, expense) %>% arrange(country, year)
table3 <- table1 %>% unite(rate, cases, population, sep = "/")
table4a <- table1 %>% select(-population,-expense) %>% spread(year, cases)
table4b <- table1 %>% select(-cases,-expense) %>% spread(year, population)
table4c <- table1 %>% select(-population,-cases) %>% spread(year, expense)
table1 
```

+ Q1. What is the dimension (number of rows and columns) of $\mathtt{table2}$ below ?
```{r eval = FALSE}
table2 <- gather(table1, key = "type", value = "value", cases, population, expense) %>% 
  arrange(country, year)
```
$$\mathrm{(a)} \text{12 and 5}; \quad \mathrm{(b)}  \text{12 and 4}; \quad  \mathrm{(c)} \text{18 and 4}; \quad \mathrm{(d)} \text{18 and 5}$$

+ Q2. How will you go from $\mathtt{table1}$ to the following table 
```{r echo = FALSE}
table4b
```
```{r eval = FALSE}
## (a)
table1 %>% select(-population,expense) %>% spread(year, cases)
## (b)
table1 %>% select(-cases,-expense) %>% spread(year, population)
## (c)
table1 %>% select(-population,-cases) %>% spread(year, expense)
## (d)
table1 %>% select(population,cases) %>% spread(year, population)
```

+ Q3. How will you go from the following tables back to $\mathtt{table1}$ ? Denote these tables as $\mathtt{table\_abbr}$, $\mathtt{table5a}$ and $\mathtt{table5b}$.
```{r echo = FALSE, message = FALSE}
table_abbr <- tribble(
  ~fullname, ~abbr, 
  "Afghanistan", "AF", 
    "Brazil", "BR", 
  "China", "CN",
  "France", "FR",
  "Spain", "SP",
)
#table_abbr
table5b <- table4b %>% left_join(table_abbr, by = c("country" = "fullname")) %>% select(-country) %>% select(abbr, everything())
table5a <- table1 %>% select(-population) 

knitr::kable(list(table_abbr, table5a, table5b))
```
```{r error = TRUE, message = FALSE, eval = FALSE}
## (a)
table_abbr %>% right_join(table5b) %>% select(-abbr) %>% 
  gather(key = "year", value = "population", `1999`, `2000`, convert = TRUE) %>%
  rename(country = fullname) %>% right_join(table5a)
## (b)
table_abbr %>% left_join(table5b) %>% select(-abbr) %>% 
  gather(key = "year", value = "population", `1999`, `2000`, convert = TRUE) %>%
  rename(country = fullname) %>% left_join(table5a)
## (c)
table_abbr %>% left_join(table5b) %>% select(-abbr) %>% 
  gather(key = "year", value = "population", `1999`, `2000`, convert = TRUE) %>%
  left_join(table5a)
## (d)
table_abbr %>% right_join(table5b) %>% select(-abbr) %>% 
  gather(key = "year", value = "population", `1999`, `2000`, convert = TRUE) %>%
  left_join(table5a, by = c("fullname" = "country"))
```

+ Q4: How will you go from **table1** to the following table (call it **table_q4**) ? Here $\mathtt{means}$ contain the sample average for each variable
```{r echo = 3:4}
colmeans <- function(df){
  sapply(select(df, -year), mean)
}
table_q4 <- table1 %>% group_by(country) %>% nest() %>% mutate(means = map(data, colmeans))
table_q4
table_q4$means[[1]]
```

```{r error = TRUE, message = FALSE, eval = FALSE}

colmeans <- function(df){
  sapply(select(df, -year), mean)
}

##(a)
table_q4 <- table1 %>% group_by(country) %>% nest() %>% 
  mutate(means = map(data, colmeans)) %>% unnest()

##(b)
table_q4 <- table1 %>% group_by(country,year) %>% nest() %>% map(data, colmeans)
##(c)
table_q4 <- table1 %>% group_by(country) %>% nest() %>% map(data, colmeans)
```

# Part 4 (Basic statistics and linear models)

For this part, read the following description. DrinkDrankDrunk\textregistered is a beverage company. The company is interested in finding the effect of choice of beverages on employees productivity. The company performs a
study in which the predictor variable $X$ is the amount of coffee (in pint) consumed by an employee per day for a duration of one month,
$Z$ is the amount of soda (in pint) consumed by an employee per day for a duration of one month, and $W$ is the
amount of energy drink (in pint) consumed by an employee per day for a duration of
one month. The response variable $Y$ is a measure of productivity of the employee for the month. The number of data points is $n = 40$. The values of the predictor variable $X$ ranges from $1$ to $5$, the values of the predictor variable $Z$ ranges from $1$ to $4$ and the values of the predictor variables $W$ ranges from $0$ to $3$. A snippet of the data is as follows

```{r echo = FALSE, message=FALSE, warning = FALSE}
set.seed(123)
n <- 40
X <- rep(c(1,2,3,4), each = 10)
Z <- rep(c(5,3,2,1), each = 10)
W <- Z/2 + rnorm(n, sd = 0.1)
Y <- 1 + 1.2*W + 0.5*X + 2*Z + rnorm(n, sd = 0.2)
Xmat <- matrix(0,n,4)
Xmat[,1] <- 1
Xmat[,2] <- W
Xmat[,3] <- X
Xmat[,4] <- Z
df <- data_frame(Y = Y, W = W, X = X, Z = Z)
df
P <- solve(t(Xmat) %*% Xmat)
```


The design matrix $\mathbf{X}$ satisfies

$$\mathbf{X} = \begin{bmatrix} 1 & X_1 & Z_1 & W_1 \\ 1 &  X_2 & Z_2 & W_2 \\ \vdots & \vdots & \vdots \\ 1 & X_n & Z_n & 
W_n \end{bmatrix}; \quad (\mathbf{X}^{T} \mathbf{X})^{-1}
= \begin{bmatrix} 12.18 & 0.28 & -2.65  & -2.15 \\ -0.28 & 3.2 & -0.06 & -1.64 
\\ -2.65 & -0.06 & 0.58 & 0.47 \\ -2.15 & -1.64 & 0.47 & 1.18   \end{bmatrix}$$ 


The following output are the $95\%$ confidence interval obtained from **R** for fitting the model
$Y_i = \beta_0 + \beta_W W_i + \beta_X X_i + \beta_Z Z_i + \epsilon_i$
to the data where the $\epsilon_i$ are assumed i.i.d. $N(0, \sigma^2)$.

```{r echo = TRUE}
mod <- lm(Y ~ W + X + Z, df)
confint(mod, level = 0.95)
```


From the above ouput, answer the following question.


Q1. What is the least square estimate for $\hat{\beta}_W$ ?

$$\mathrm{(a)} `r coef(mod)[2]`; \quad \mathrm{(b)} `r coef(mod)[2] + 0.5`; \quad \mathrm{(c)} `r coef(mod)[2] - 0.5`; \quad \mathrm{(d)} `r coef(mod)[2]/2`$$
Q2. What is a reasonable estimate of $\sigma^2$ for this data (assuming the posited model is "correct") ? You can assume that the $97.5\%$ quantile for the appropriate Student-$t$ distribution is $2$.
$$\mathrm{(a)} `r summary(mod)$sigma^2 + 1`; \quad \mathrm{(b)} `r summary(mod)$sigma`; \quad \mathrm{(c)} `r summary(mod)$sigma^2`; \quad \mathrm{(d)} `r summary(mod)$sigma^2/2`$$
Q3. What is a reasonable **prediction** interval for $Y_*$ when $\boldsymbol{x}_* = (W_*, X_*, Z_*) = (2,3,4)$ ?
(This is a fill-in a blank question. You don't have to provide the exact numbers but you need to demonstrate that you know how to obtain the answer from the output given above).

Q4. Can you conclude, from the above output, that (assuming the model is true), there is scant evidence supporting the hypothesis that 
$\beta_X = 0.65$ and $\beta_Z = 1$ **simultaneously** (with $\beta_0$ and $\beta_W$ arbitrary) ? Why or why not ? 


# Part 5 (Visualizing data)

For this part, use the following dataset on highway efficiency of cars.
```{r}
mpg
```

Q1. Which code will reproduce the following figure ?
```{r eval = TRUE, echo = FALSE}
library(ggplot2)
mpg %>% ggplot(aes(x = reorder(drv,hwy), y = hwy, color = factor(year))) + 
  geom_point() + geom_boxplot(aes(group = drv)) + xlab("drv") + ylab("hwy")
```

```{r eval = FALSE}
## (a)
mpg %>% ggplot(aes(x = drv, y = hwy, color = factor(year))) + geom_point() + 
  geom_boxplot(aes(group = drv))

## (b)
mpg %>% ggplot(aes(x = drv, y = hwy, color = factor(year))) + 
  geom_point() + geom_boxplot()

## (c)
mpg %>% ggplot(aes(x = reorder(drv,hwy), y = hwy, color = factor(year))) + 
  geom_point() + geom_boxplot()

## (d)
mpg %>% ggplot(aes(x = reorder(drv,hwy), y = hwy, color = factor(year))) + 
  geom_point() + geom_boxplot(aes(group = drv))
```


Q2. Which code will reproduce the following figure ?
```{r eval = TRUE, echo = FALSE}
mpg %>% ggplot(aes(x = displ,  y = cty, group = factor(year))) + geom_point() + 
  stat_smooth(method = "lm", se = FALSE) + facet_wrap( ~ drv)
```

```{r eval = FALSE, error = TRUE, message = FALSE}
## (a)
mpg %>% ggplot(aes(x = displ,  y = cty, color = year)) + geom_point() + 
  stat_smooth(method = "lm", se = FALSE) 

## (b)
mpg %>% ggplot(aes(x = displ,  y = cty, group = factor(year))) + geom_point() + 
  stat_smooth(method = "lm", se = FALSE) + facet_wrap(~ drv)

## (c)
mpg %>% ggplot(aes(x = displ,  y = cty, color = year)) + geom_point() + 
  stat_smooth() + facet_wrap(~ drv)

## (d)
mpg %>% ggplot(aes(x = displ,  y = cty, color = year)) + geom_point() + 
  stat_smooth(aes(group = drv))
```