---
title: "Introduction to Data Science"
subtitle: "Text processing with R"
date: "Fall 2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    #css: ["default","metropolis","metropolis-fonts","animate.css"]
    css: ["xaringan-themer.css","metropolis-fonts"]
    nature:
      highlightStyle: solarized-light
      highlightLines: false
      highlightSpans: false
      countIncrementalSlides: false
    # df_print: tibble
    # css: [default, metropolis, metropolis-fonts]
    # nature:
    #   highlightStyle: zenburn
    #   highlightLines: true
    #   countIncrementalSlides: false
--- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.retina = 3, fig.asp = 0.6, fig.align = 'center', out.width = "120%",message = FALSE)
options(htmltools.dir.version = FALSE, digits = 3, knitr.table.format = "html",tibble.print_min=6, tibble.print_max = 6, tibble.width=70)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
xaringanExtra::use_tile_view()
xaringanExtra::use_scribble()
xaringanExtra::use_extra_styles(hover_code_line = TRUE)
xaringanExtra::use_search(show_icon = TRUE)
xaringanExtra::use_tachyons()
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
style_duo_accent(primary_color = "#035AA6", secondary_color = "#03A696",
   #title_slide_background_color = "#FFFFFF",
   #title_slide_text_color = "#006747",
   link_color = "#03A696",
   header_font_google = google_font("Josefin Sans"),
   title_slide_background_image = "belltower2.jpg",
   #title_slide_background_size = "600px",
   #title_slide_background_position = "bottom",
   text_font_google   = google_font("Montserrat", "300", "300i"),
   code_font_size = "0.8rem",
   code_font_family = "Fira Code",
   code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```
# Additional Readings

+ J. Silge and D. Robinson [Text Mining with R: A tidy approach](https://www.tidytextmining.com/)

+ Chapter 26 of R. Irizarry [Introduction to Data Science](https://rafalab.github.io/dsbook/text-mining.html)

+ Chapter 19 of B. Baumer, D. Kaplan and N. Horton [Modern Data Science with R, 2nd edition](https://mdsr-book.github.io/mdsr2e/ch-text.html)

+ M. Jockers and R. Thalken [Text analysis with R](https://catalog.lib.ncsu.edu/catalog/NCSU4839893)

+ E. Hvitfeld and J. Silge [Supervised Machine Learning for Text Analysis in R](https://smltar.com/)

---
# Text data and tokenization.

Let us first consider the following two stanzas from a poem by [Oscar Wilde](https://en.wikipedia.org/wiki/Oscar_Wilde)

```{r echo = -c(1:2)}
library(stringr)
text <- str_c("Tread lightly, she is near\r\n",
"\tUnder the snow\r\n",
"Speak gently, she can hear\r\n",
"\tThe daisies grow.\r\n",
"All her bright golden hair\r\n",
"\tTarnished with rust\r\n",
"She that was young and fair\r\n",
"\tFallen to dust.", collapse = "")
cat(text)
```
--
The raw text looks like 
```{r}
text
```

---
class: clear, middle

In text mining, the first thing to do is generally to **tokenize** a given text into tokens such as letters, words, and more general [n-grams](https://en.wikipedia.org/wiki/N-gram). 

We will use the [unnest_tokens](https://www.rdocumentation.org/packages/tidytext/versions/0.3.1/topics/unnest_tokens) function from the [tidytext](https://www.rdocumentation.org/packages/tidytext/versions/0.3.1) package.

```{r}
library(tidytext) ## 
library(stringr)
library(tibble)
text_lines <- text %>% str_split(pattern = boundary("sentence")) 
text_lines
text_df <- tibble(lines = 1:length(text_lines[[1]]), text = text_lines[[1]])
text_df
```

---
class: clear, middle
We can now tokenize the above two stanzas into words.
```{r}
tokens <- text_df %>% unnest_tokens(words, text)
tokens
```
Other tokenization are also possible, e.g.,
```{r}
text_df %>% unnest_tokens(ngrams, text, token = "ngrams", n = 3)
```

---
# Jane Austen and tokenization.
The following is an abridged presentation of Chapter 1 of [Text Mining with R](https://www.tidytextmining.com/).
```{r}
library(janeaustenr) ## The package contains the text of 6 Jane Austen's novels.
library(dplyr)
austen_books()
austen_books() %>% dplyr::select(book) %>% unique() %>% pull()
```

---
class: clear
Let us first add the line number and chapter number to each book.
```{r}
library(janeaustenr)
library(dplyr)
library(stringr)

original_books <- austen_books() %>%
  group_by(book) %>%
  mutate(linenumber = row_number(),
         chapter = cumsum(str_detect(text, 
                                     regex("^chapter [\\divxlc]",
                                           ignore_case = TRUE)))) %>%
  ungroup()

original_books
```
---
class: clear
We next tokenize the text into words and remove the common stop words.
```{r}
data(stop_words) ## Part of the tidytext package.
jausten_tokenized <- original_books %>% unnest_tokens(word, text) %>% 
  anti_join(stop_words)
jausten_tokenized
```
---
Given this tidy representation, it is easy to do some data exploration. For example, what are the most common words ?
```{r}
words_freq <- jausten_tokenized %>% count(word, sort = TRUE)
words_freq
```
Which book is most verbose ?
```{r}
jausten_tokenized %>% group_by(book) %>% 
  summarize(no_words = n()) %>% 
  arrange(desc(no_words))
```
---
Maybe a words cloud would be nice ?
```{r out.width="90%", fig.align = "center"}
library(wordcloud)
wordcloud(words = words_freq$word, freq = words_freq$n, min.freq = 500, 
          colors = brewer.pal(8, "Dark2"))
```

---
Is a word cloud better than a bar plot or a dot plot ?
```{r fig.show = "hold", out.width="90%"}
library(ggplot2)
p <- words_freq %>% filter(n > 500) %>% mutate(word = reorder(word,n)) %>% 
  ggplot(aes(n, word)) + labs(y = NULL)
p1 <- p + geom_col()
p2 <- p + geom_point()
gridExtra::grid.arrange(p1,p2,nrow = 1)

```
---
# Battle of the authors
Let us now compare the word distributions of Jane Austen against other authors like Herberg G. Wells and the Brontës sisters.

```{r gutenberg1, cache = TRUE}
## First download the data.
library(gutenbergr)
hgwells <- gutenberg_download(c(35, 36, 5230, 159))
tidy_hgwells <- hgwells %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words)
bronte <- gutenberg_download(c(1260, 768, 969, 9182, 767))
tidy_bronte <- bronte %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words)
```

---
```{r}
glimpse(tidy_hgwells)
glimpse(tidy_bronte)
tidy_austen <- jausten_tokenized
## Let us now create a data frame with these words and their authors
frequency <- bind_rows(mutate(tidy_bronte, author = "Bronte sisters"),
                       mutate(tidy_hgwells, author = "H.G. Wells"),
                       mutate(tidy_austen, author = "Jane Austen")) %>%
  mutate(word = str_extract(word, "[a-z']+"))
```

---
We now see, for each author, how likely a specific word is used
```{r}
library(tidyr)
frequency <- frequency %>% group_by(author, word) %>% summarise(n = n()) %>%
  mutate(proportion = n/sum(n))%>%  ## grouped mutate! 
  select(-n) %>% pivot_wider(names_from = author, values_from = proportion)
frequency
```

---
We can now visualize the distribution of the words used by the Brontes sisters and H. G. Wells, when compared to the Jane Austen baseline. 
```{r warning = FALSE}
p1 <- ggplot(frequency, aes(x = `Bronte sisters`, y = `Jane Austen`), 
            color = abs(`Jane Austen` - `Bronte sisters`)) +
  geom_abline(color = "gray40", lty = 2) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = scales::percent_format()) +
  scale_y_log10(labels = scales::percent_format()) +
  scale_color_gradient(limits = c(0, 0.001), 
                       low = "darkslategray4", high = "gray75") +
  theme(legend.position="none") +
  labs(y = "Jane Austen", x = NULL) + ggtitle("Bronte Sisters")
p2 <- ggplot(frequency, aes(x = `H.G. Wells`, y = `Jane Austen`), 
            color = abs(`Jane Austen` - `H.G. Wells`)) +
  geom_abline(color = "gray40", lty = 2) +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
  geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = scales::percent_format()) +
  scale_y_log10(labels = scales::percent_format()) +
  scale_color_gradient(limits = c(0, 0.001), 
                       low = "darkslategray4", high = "gray75") +
  theme(legend.position="none") +
  labs(y = "Jane Austen", x = NULL) + ggtitle("H.G. Wells")
```
---
```{r out.width="70%", fig.asp = 1, warning = FALSE}
p1
```
---
```{r out.width="70%", fig.asp = 1, warning = FALSE}
p2
```